### 2022-05-26 Manuscript/Manuscript.tex
Update line numbers

\end{flushleft}
\newpage
\section*{
Abstract
}
The skills and knowledge needed to recognize and classify taxa is becoming increasingly scarce in the scientific community. At the same time, it is clear that these skills are strongly needed in biodiversity monitoring for management and conservation, especially when carried out by citizen scientists. Formalizing the required knowledge in the form of digital identification keys is one way of making such knowledge more available for professional and amateur observers of biodiversity. In this paper we describe Clavis, a modern open format for capturing knowledge required for taxon identification through digital keys, allowing for a level of detail beyond that of any current key format. We exemplify each concept using Pokémon as a fictional taxonomic group.
\section*{
Introduction
}
Distinguishing biological taxa from one another is a necessity in biodiversity monitoring. As species are going extinct at an unprecedented rate~\cite{Ceballos2015, Johnson2017}, we need to monitor as much of nature as we can to identify population statuses and trends. Meanwhile, research and management is currently facing a ``taxonomic impediment'', where taxonomic knowledge is gradually disappearing from the scientific community~\cite{Engel2021}.

@@

\textit{\textbf{Example}: Pikachu is always an electric type. See lines 264 - 270 in the \nameref{S2_Appendix} for the corresponding JSON.}

\textit{\textbf{Example}: Pikachu is never blue. See lines 271 - 277 in the \nameref{S2_Appendix} for the corresponding JSON.}

\textit{\textbf{Example}: Pikachu has a double-lobed tail in 50\% of the cases. See lines 278 - 291 in the \nameref{S2_Appendix} for the corresponding JSON.}

@@

\textit{\textbf{Example}: The rain form of Castform has a higher frequency in the notoriously rainy Norwegian city of Bergen. See lines 302 - 347 in the \nameref{S2_Appendix} for the corresponding JSON.}
\subsection*{
Logical premises
}
Not all characters are meaningful in every context, and it is sometimes desirable to be able to specify conditions that must be met before a character is shown to the user. Such a condition may be that the user has given a certain answer to some other character in the key first. To this end, characters can have a logical premise, specifying which facts have to be established for it to be relevant. Logical premises can also relate several facts combined, including numerical values, and consist of strings using unary and binary operators in JavaScript notation.

@@

\textit{\textbf{Example}: The weight of a Pokémon varies, both between individuals within a species and between species. To distinguish Pokémon based on their weight, this is added as a numerical character, where weights can be specified in whole kilograms within a given range. A statement can then be included to register the possible weight range of a given taxon. See lines 249 - 261 (character) and lines 292 - 301 (statement) in the \nameref{S2_Appendix} for the corresponding JSON.}
\subsection*{
External services
}
Various units in a key may be subject to changes that are ideally managed outside the key, in designated centralized systems. Taxon names can change, as can urls to supplemental information, media files, geographic ranges etc. Other parameters, such as the probability distribution of a numerical character for a given taxon, may be too detailed for direct inclusion in a key. It is best practice to not duplicate such resources, but to harvest these via an Application Programming Interface (API) or other service interface. To facilitate this, Clavis allows the specification of external services, where documentation for use of the service can be linked. Various units in the key, such as taxa, can refer to such external services through one or more externalResources, connecting the service and the relevant external id to the taxon at hand.

@@

\textit{\textbf{Example}: To allow for features such as the retrieval of the name of a Pokémon in different languages, one can use a Wikidata query. See lines 435 - 439 (service) and lines 47 - 50 (externalResource) in the \nameref{S2_Appendix} for the corresponding JSON.}

\textit{\textbf{Example}: A hypothetical API returns the probability for a provided taxon, given its location and size. See lines 440 - 446 in the \nameref{S2_Appendix} for the corresponding JSON.}
\subsection*{
Required expertise
}
Some characters are harder to evaluate than others, and may even require special equipment. To warn and assist users, a key can contain userRequirements, describing such required skills or tools. These requirements can be connected to characters so that the user might filter out characters requiring skills or equipment they do not have, be presented with additional info to complete the task, or simply be warned.

\textit{\textbf{Example}: the weight of a Pokémon can be a useful characteristic as their weights vary between types. One can only weigh a Pokémon once it has been caught, however. The user needs to be aware of this so that in situations where a Pokémon cannot be caught (e.g. when identifying from a picture or being out of Pokéballs) such questions can be ignored. One can also add a guide on how to catch a Pokémon in order to weigh it. See lines 249 - 261 (character) and lines 420 - 432 (userRequirement) in the \nameref{S2_Appendix} for the corresponding JSON.}

@@

\textit{\textbf{Example}: The states in the tail character of Pikachu refer to images illustrating the different tail shapes. See lines 207 - 218 (states) and lines 449 - 498 (mediaElements) in the \nameref{S2_Appendix} for the corresponding JSON.}
\subsection*{
Descriptions
}
There are many aspects of an identification key where a more elaborate description is desirable or even required. To this end, many elements can have a short and/or extended description in valid markdown notation, as well as a url to an online description.

\textit{\textbf{Example}: A description can be used to explain to the user how to catch a Pokémon where relevant, by including a short description and a link to a page with more information. See lines 425 - 431 in the \nameref{S2_Appendix} for the corresponding JSON.}
\subsection*{
Followup keys
}
Once a key has identified a taxon as far as it is intended to, i.e. when it has narrowed it down to an endpoint taxon, the user is presented with the result. It may however be possible to further determine the result. If a key exists somewhere that can help with this, it can be referred to as a followup key from the relevant taxon. It can be either a url, or a reference to an (external) service. This feature can also be used to split keys into several smaller keys that refer to one another through this mechanism.

>>>>

\end{flushleft}
\newpage
\section*{
Abstract
}
The skills and knowledge needed to recognize and classify taxa is becoming increasingly scarce in the scientific community. At the same time, it is clear that these skills are strongly needed in biodiversity monitoring for management and conservation, especially when carried out by citizen scientists. Formalizing the required knowledge in the form of digital identification keys is one way of making such knowledge more available for professional and amateur observers of biodiversity. In this paper we describe Clavis, a modern open format for capturing knowledge required for taxon identification through digital keys, allowing for a level of detail beyond that of any current key format. We exemplify each concept using Pokémon as a fictional taxonomic group.

\section*{
Introduction
}
Distinguishing biological taxa from one another is a necessity in biodiversity monitoring. As species are going extinct at an unprecedented rate~\cite{Ceballos2015, Johnson2017}, we need to monitor as much of nature as we can to identify population statuses and trends. Meanwhile, research and management is currently facing a ``taxonomic impediment'', where taxonomic knowledge is gradually disappearing from the scientific community~\cite{Engel2021}.

@@

\textit{\textbf{Example}: Pikachu is always an electric type. See lines 261 - 267 in the \nameref{S2_Appendix} for the corresponding JSON.}

\textit{\textbf{Example}: Pikachu is never blue. See lines 268 - 274 in the \nameref{S2_Appendix} for the corresponding JSON.}

\textit{\textbf{Example}: Pikachu has a double-lobed tail in 50\% of the cases. See lines 275 - 281 in the \nameref{S2_Appendix} for the corresponding JSON.}

@@

\textit{\textbf{Example}: The rain form of Castform has a higher frequency in the notoriously rainy Norwegian city of Bergen. See lines 299 - 344 in the \nameref{S2_Appendix} for the corresponding JSON.}
\subsection*{
Logical premises
}
Not all characters are meaningful in every context, and it is sometimes desirable to be able to specify conditions that must be met before a character is shown to the user. Such a condition may be that the user has given a certain answer to some other character in the key first. To this end, characters can have a logical premise, specifying which facts have to be established for it to be relevant. Logical premises can also relate several facts combined, including numerical values, and consist of strings using unary and binary operators in JavaScript notation.

@@

\textit{\textbf{Example}: The weight of a Pokémon varies, both between individuals within a species and between species. To distinguish Pokémon based on their weight, this is added as a numerical character, where weights can be specified in whole kilograms within a given range. A statement can then be included to register the possible weight range of a given taxon. See lines 249 - 258 (character) and lines 289 - 298 (statement) in the \nameref{S2_Appendix} for the corresponding JSON.}
\subsection*{
External services
}
Various units in a key may be subject to changes that are ideally managed outside the key, in designated centralized systems. Taxon names can change, as can urls to supplemental information, media files, geographic ranges etc. Other parameters, such as the probability distribution of a numerical character for a given taxon, may be too detailed for direct inclusion in a key. It is best practice to not duplicate such resources, but to harvest these via an Application Programming Interface (API) or other service interface. To facilitate this, Clavis allows the specification of external services, where documentation for use of the service can be linked. Various units in the key, such as taxa, can refer to such external services through one or more externalResources, connecting the service and the relevant external id to the taxon at hand.

@@

\textit{\textbf{Example}: To allow for features such as the retrieval of the name of a Pokémon in different languages, one can use a Wikidata query. See lines 432 - 436 (service) and lines 47 - 50 (externalResource) in the \nameref{S2_Appendix} for the corresponding JSON.}

\textit{\textbf{Example}: A hypothetical API returns the probability for a provided taxon, given its location and size. See lines 437 - 443 in the \nameref{S2_Appendix} for the corresponding JSON.}
\subsection*{
Required expertise
}
Some characters are harder to evaluate than others, and may even require special equipment. To warn and assist users, a key can contain userRequirements, describing such required skills or tools. These requirements can be connected to characters so that the user might filter out characters requiring skills or equipment they do not have, be presented with additional info to complete the task, or simply be warned.

\textit{\textbf{Example}: the weight of a Pokémon can be a useful characteristic as their weights vary between types. One can only weigh a Pokémon once it has been caught, however. The user needs to be aware of this so that in situations where a Pokémon cannot be caught (e.g. when identifying from a picture or being out of Pokéballs) such questions can be ignored. One can also add a guide on how to catch a Pokémon in order to weigh it. See lines 249 - 258 (character) and lines 417 - 429 (userRequirement) in the \nameref{S2_Appendix} for the corresponding JSON.}

@@

\textit{\textbf{Example}: The states in the tail character of Pikachu refer to images illustrating the different tail shapes. See lines 207 - 218 (states) and lines 446 - 496 (mediaElements) in the \nameref{S2_Appendix} for the corresponding JSON.}
\subsection*{
Descriptions
}
There are many aspects of an identification key where a more elaborate description is desirable or even required. To this end, many elements can have a short and/or extended description in valid markdown notation, as well as a url to an online description.

\textit{\textbf{Example}: A description can be used to explain to the user how to catch a Pokémon where relevant, by including a short description and a link to a page with more information. See lines 422 - 429 in the \nameref{S2_Appendix} for the corresponding JSON.}
\subsection*{
Followup keys
}
Once a key has identified a taxon as far as it is intended to, i.e. when it has narrowed it down to an endpoint taxon, the user is presented with the result. It may however be possible to further determine the result. If a key exists somewhere that can help with this, it can be referred to as a followup key from the relevant taxon. It can be either a url, or a reference to an (external) service. This feature can also be used to split keys into several smaller keys that refer to one another through this mechanism.
